\section{Vedic, Brahmanic and Jain Scriptures}

Various authors have already noticed the similarities of \textit{Vinaya} terminology between the Buddhist and Jain Orders.\footnote{\cite{maes2016} page 9 footnotes 26–28, also \cite{sujato2009} and \cite{zwilling}.} In the Buddhist \textit{Suttas} we find many examples of discussions between the two groups and after the Buddha passed away, when the community found itself without a leader to make decisions, such discussions would certainly have had an impact on the Buddhist \textit{Vinaya}. Both groups would have also used vocabulary that was already in existence at that time. It is therefore important to first have a look at what sex and gender meant in the larger context of the society in pre-Buddhist India and how this understanding developed within the Jain Order.

\subsection{Emergence of the Third Sex/Gender}
Both the Buddhist and Vedic texts have origin stories that describe the emergence of sex and gender (and thus human beings) as a result of desire.\footnote{\cite{zwilling2000} page 101 and \cite{artinger} pages 7–15.} Before this emergence of sex and gender, beings were androgynous, hermaphroditic or without any kind of sexual characteristics.

In the Vedic myths and legends, we frequently find the recurrent theme of a man turning into a woman or androgyn or of being both a mother and a father.\footnote{F.i. \textit{Mahābārata Anuśāsanaparvan} (MBh) 13.12 and \textit{Rāmāyaṇam} (Rām) 7.78–79. See also \cite{goldman} pages 379–380 and \cite{cassell} page 68.} Robert Goldman\footnote{Robert \cite{goldman} gives an excellent account of the myths that formed the notions of sex and gender in ancient India and the psychological purpose of these myths. I will refer to some of these stories throughout this paper.} and others point out that the function of these myths and legends is to confront deep anxieties and fears associated with the complex and problematic issues involving sex, gender, sexuality, power, hierarchy and subordination. We see literary representations of these anxieties in all patriarchal societies, expressing the deeply ambivalent attitude towards women and women's sexuality. On the one hand, women are depicted as pure and nurturing as long as they are controlled within the constraints of kinship, but outside such a regulated environment they are seen as dangerous and destructive to men. Through such projective devices of men onto women, male-dominated cultures have been able to establish a hegemonic ideology of gender.\footnote{See \cite{sujato2011} for an extensive work on the role of the male/female relationship portrayed in mythology in Buddhism and more specifically with regards to women's ordination.} We see that `transgenderism' is a recurring theme in these myths and legends, derived from these anxieties and attitudes towards sex and gender in a society where the `male' sex and sexuality is seen as superior.

We can trace the emergence of the concept of a third sex/gender back to the late Vedic period (800–600 BCE).\footnote{Throughout this paper I have tried to maintain a distinction between `sex'--the anatomy of an individual's reproductive system--and `gender'--either social roles or personal identification based on an internal awareness--that we use today, but this has proven rather difficult because in the ancient texts this disctinction is significantly blurred.} The term \textit{napuṁsaka} was originally an umbrella term used to denote men who were impotent, who had a female gender-expression or who dressed in traditional women's clothing.\footnote{\cite{zwilling} page 362 and \cite{zwilling2000} page 105.} Literally the term means `not a male' i.e. men who did not conform to gender-role expectations.

It is accepted by many scholars that the \textit{hijras} of India are the contemporary representatives of these \textit{napuṁsakas} and they even refer to themselves as `not a male' or `neither male nor female'.\footnote{\cite{zwilling} page 363, \cite{goldman} page 388 and \cite{wendy}. A good study on the \textit{hijras} is provided by Serena \cite{nanda}.} Although the \textit{hijras} have been subject to influences from the Muslim period from the 12\textsuperscript{th} to the 16\textsuperscript{th} Century as well as from Christian and more modern Western ideas, by studying them we get a glimpse of what the lives of the \textit{napuṁsakas} must have been like in Vedic times. 

The \textit{hijras} are a representation of the god Śiva in their androgynous form of Ardhanārīśwara from Vedic mythology (See Figure \ref{siva}). They enact the religious myths and make them come alive in the rituals, songs and dances they perform. They are viewed as vehicles of the divine power of the Mother Goddess, which transforms their impotence into the power of creation. According to sources from the beginning of he 20\textsuperscript{th} Century, potential candidates for initiation as a \textit{hijra} (which includes emasculation) had to be born impotent.\footnote{The 19\textsuperscript{th} and early 20\textsuperscript{th} Century sources are important because at that time the influence of globalization and the influx of Western ideas on the lives of the \textit{hijras} was less prevalent than it is today. See \cite{ibbetson} pages 332, \cite{shah} page 1325 and \cite{bhimbhai}. Some 19\textsuperscript{th} Century accounts report that impotence was an essential qualification for admission into the hijra community and that a newcomer initiated into the community was on probation for as long as a year. During this time his impotence was carefully tested, sometimes by making the person sleep four nights `with a prostitute'. Only after impotence was established would the newcomer be permitted to undergo the emasculation operation and become a full member of the community. \cite{preston}.} Their emasculated body is not the only characteristic of a \textit{hijra}, but also their physiology and their sexual capacities, feelings, preferences and behaviors. Although dressing like a woman is part of a \textit{hijra}, they are also quite different from a transgender person; they are the religious embodiment of the deities.\footnote{\cite{nanda}.}

Next to \textit{hijras} there are a number of other religious sects in India where males impersonate the legends, imitate what is seen as feminine dress and behavior, including menstruation, and also engage in sexual acts with men as acts of devotion. Castration may also be part of this like f.i. in the \textit{Sakhibhāva} sect that worships Viṣṇu.\footnote{\cite{nanda} page 21.} Various \textit{sādhus}, like Ramakrishna, identify themselves with women for periods of time as part of their religious life.\footnote{Robert \cite{goldman} pages 384–391 discusses various sects with these practices.} 

The adoption of the word \textit{napuṁsaka} as a grammatical third gender\footnote{\textit{Śatapatha Brāhmaṇa} (ŚB) 10.5.1.2–3.} in the 6\textsuperscript{th} Century BCE seems to have prompted a significant shift in meaning, because now the \textit{napuṁsaka} was interpreted as meaning `neither male nor female'. This resulted in the previously mentioned `not males' being regarded as people with ambiguous sex or gender.\footnote{\cite{zwilling2000} page 362.} The individuals that the word \textit{napuṁsaka} originally referred to were however all males, just not conforming to gender-role expectations. The fact that the word \textit{liṅga} (sex)\footnote{The original meaning of \textit{liṅga} is `characteristic mark or sign' (\textit{Nirukta} 1.17) but it later starts to mean `sexual characteristic'. See also chapter \ref{linga} for a detailed description of \textit{liṅga}.} was adopted as the technical term for grammatical gender created a lot of confusion because now the same word was used to express natural sex and grammatical gender.\footnote{\cite{zwilling} page 365.} The third grammatical gender for humans became equivalent to the `third sex' and comprised of anybody who did not neatly fit into the categories of `male' and `female'.

% The fact that Sanskrit is a gendered language forced people to assign a gender to all objects including all living beings and humans. Gender was seen as a property belonging to objects and objects are gendered by the presence or absence of certain defining characteristics or \textit{liṅga} (sex).\footnote{The original meaning of \textit{liṅga} is `characteristic mark or sign' (Nirukta 1.17) but it later starts to mean `sexual characteristic'. See also chapter \ref{linga} for a detailed description of \textit{liṅga}.} This led to speculations on the characteristics by which a person can be identified as belonging to one of the three grammatical genders in the context of examening the relationship between sex and grammatical gender. The third gender (\textit{napuṁsaka}) was basically a class for things that were neither male nor female in nature. The adoption of the word \textit{liṅga} as a technical term for grammatical gender created an intimate connection between sex and grammatical gender that caused much confusion\footnote{\cite{zwilling} page 365. Other languages, like the Uralic languages, do not have gender in language. Such gender-less languages exclude many possibilities for reinforcement of gender-related stereotypes as they do not place objects (and thus people) in boxes.} 

Just after the late Vedic period we see that a set of terms relating to the class of \textit{napuṁsakas} emerged like \textit{klība} (sexually defective man\footnote{As pointed out by \cite{zwilling} page 363, the nature of the \textit{klība} is suggested by the \textit{Bṛhadāraṇyaka Upaniṣad} 6.1.12 and is the result of the destruction of the penis as in ŚB 1.4.3.19.}), \textit{ṣaṇḍha}\footnote{The Monier Monier-Williams dictionary gives the translation `eunuch, hermaphrodite'.} and \textit{paṇḍaka} (`impotent', or `sterile'\footnote{\textit{Atharva Veda} (AV) 8.6.7, 11.16. The etymology of \textit{paṇḍaka} is unknown but cf. \textit{baṇḍa} at AV 7.65.3 is glossed by the commentator as \textit{nirvīrya} (`impotent, powerless, unmanly' (Monier-Williams, Sir M. (1988) dictionary). Albrecht \cite{wezler} has suggested that \textit{paṇḍa} and \textit{paṇḍaka} be regarded as ultimately derived from \textit{*apa}+ \textit{āṇḍā}, thus: ``one who has no testicles (anymore),'' i.e. a eunuch. Allan \cite{bomhard} points out that there is a range of translations and interpretations that can apply to the \textit{paṇḍaka}. He believes that the word could be a loan-word from the Dravidian \textit{peṇṭan, peṇṭakan, peṇṭakam}, which can mean both hermaphrodite and eunuch.}). All of these types were associated with cross-dressing, singing and dancing,\footnote{The term \textit{klība} surfaces in the \textit{Mahābhārata} in the myth of Arjuna, who refuses the advances of a nymph who curses him to become a \textit{klība}. Arjuna takes the name Brihannala, dresses in women's clothes and gains access to the princess Uttarā for whom she becomes a performer in the arts of music, singing and dancing. See \cite{cassell} for details.} as we have also seen in the above description of the contemporary \textit{hijras}. With the word \textit{napuṁsaka} having gained a much broader meaning, it seems likely that these new sub-categories represent different names for the original meaning of `not males', although not clearly defined and overlapping in meaning.

Serena Nanda\footnote{Serena \cite{nanda} quotes Bullough pages 21–22.} mentions the practices and beliefs in Tantric Hinduism which started emerging in India from the middle of the 1\textsuperscript{st} millennium CE:

\begin{quote}
... there was a third sex, which itself was divided into four categories: the male eunuch, called the `waterless' because he had desiccated testes; the `testicle voided', so called because he had been castrated; the hermaphrodite; and the `not woman', or female eunuch (which usually refers to a woman who does not menstruate). Those who were more feminine (whether males or females) wore false breasts and imitated the voice, gestures, dress, delicacy, and timidity of women. All these categories of persons had the function of providing alternative techniques of sexual gratification, some of which are mentioned in the classical Hindu sex manual, the \textit{Kāmasūtra}.
\end{quote}

This shows that at that time the `third sex', the \textit{napuṁsakas}, were not only defined by their primary sexual characteristics that made them neither completely `male' or `female', but they were also prescribed gender characteristics like dress and behavior as well as a certain religious sexual role in society. I have been unable to identify the Sanskrit words that the four categories mentioned refer to, but it seems likely they represent at least the \textit{paṇḍaka}, the \textit{klība} or \textit{ṣaṇḍha} and possibly the female \textit{paṇḍaka}. The \textit{Kāmasūtra} itself only uses the word \textit{tṛtīyāprakṛtī} which literally means `third sex'.\footnote{At the beginning of the Common Era the term \textit{tṛtīyāprakṛtī} was introduced, possibly among the schools of traditional medicine, to indicate a true `third sex'. This term is usually used as a synonym for \textit{napuṁsaka} and only appears in a few times in the later Brahmanic Sastra collections (See \cite{zwilling} page 362). The Pali equivalent of the word is \textit{tatiya pakati}, which appears a few times in later Buddhist \textit{Abhidhamma} sub-commentaries.} The characteristics of the `hermaphrodite' and which Sanskrit term is used here remains unclear. I will discuss this in the next chapters.

\subsection{Sex and Gender in the Jain Order}
Just like in Buddhism, the Jain Order had a strong interest in controlling the sexuality of its monastics. Jain monastics practiced celibacy,\footnote{Celibacy is the state of voluntarily being unmarried, sexually abstinent, or both, usually for religious reasons. In this paper I will use the word in the meaning of sexually abstinent.} and the monks were mostly naked ascetics. The prestige and power of the Order depended to a large extend on public opinion and therefore on the purity of their behavior, as well as their external appearance. The `third sex' was therefore the subject of a very lengthy debate within the Order. 

In addition to these practical considerations, there was also a debate within the Jain community as to whether women can attain spiritual liberation because the monks felt it was improper for them to go naked. Eventually it was this dispute that led to the schism between the two major Jain Orders.\footnote{See Paul \cite{dudas} for an extensive account of the schism. The two main sects of Jainism, the \textit{Digambara} and the \textit{Śvētāmbara} sects, likely started forming about the 3\textsuperscript{rd} Century BCE and the schism was complete by about the 5\textsuperscript{th} Century CE.} This controversy hinged on the identification of the signs to designate somebody as a woman, which logically also led to the examination of what is male, and `neither male nor female'. 

The speculations and discussions that followed focused on the characteristics necessary to identify a person as belonging to one of the three groups. The \textit{paṇḍaka}, \textit{klība} and \textit{keśavan} (long-haired male)\footnote{Apparently long hair was seen as a sign of a woman.} were recognized as males, but their gender-role nonconformity assimilated them to females, so not `real males' and therefore still \textit{napuṁsakas}.

This discussion was influenced by the Brahmanical views at that time concerning the essential markers for sex/gender assignment (\textit{liṅga}).\footnote{I will discuss the meaning of this term in more detail in chapter \ref{linga}.} By the 3\textsuperscript{rd} Century BCE two views had developed to define \textit{liṅga}.

\begin{enumerate}
 \item The first view went from the premise that \textit{liṅga} was defined by what one perceived as a man, woman or neither based on the presence or absence of primary or secondary characteristics.\footnote{\textit{Mahābhāṣya} 4.1.3: Q: ``What is it that people see when they decide, this is a woman, this is a man, this is neither woman nor man?'', A: ``That person who has breasts and long hair is a woman; that person who is hairy all over is a man; that person who is different from either when those characteristics are absent, is \textit{napuṁsaka}.''}
 \item The second view is that \textit{liṅga} assignment has to do with the ability to procreate or conceive. 
\end{enumerate}

Both these Brahmanical views were rejected by the Jains as being inadequate to determine sex/gender. Paul \cite{dundas} describes how the Jains developed a system to define sex/gender as a combination of sex, behavior, physical characteristics and also the underlying sexuality and feelings. The Jains came up with their own term \textit{veda} to describe these characteristics.\footnote{This move is rather remarkable because for the Brahmins \textit{veda} meant their sacred knowledge and scriptures. But it is not unprecedented because the Jains often used existing words and gave them new meaning. In the Buddhist \textit{Suttas} we also find instances where the Buddhists use different terms for the same things as the Jains. F.i. \textit{Majjhima Nikāya} 56 recounts a discussion between the Buddha and the Jain ascetic Tapassī in which the ascetic mentions that their leader Nigaṇṭha Nātaputta (i.e. Mahāvīra) doesn't speak in terms of ‘deeds', like the Buddha, but uses the term ‘rods'. See also \cite{zwilling} note 34.} This conception of sexuality most likely predates the schism between the two major Jain sects but was not part of the earliest Jain doctrine. This concept appears frequently in the later canonical Jain texts but is also mentioned once in the early Jain literature where male sexuality is explained as sexual desire for women and female sexuality is explained as sexual desire for men.\footnote{See \textit{Viyāha} 2.5.1.} The sexuality of the \textit{napuṁsaka} is not clearly defined in the earlier texts but is seen as a threat to the chastity of monks.\footnote{See \textit{Ācārāṅga Sūtra} (English translation by \cite{jacobi}) p.220: monks are warned that a danger of drunkenness is seduction by a woman or a \textit{klība}. Page 285: sleeping places frequented by women or \textit{paṇḍakas} are to be avoided.}

Zwilling and Sweet\footnote{\cite{zwilling} page 368.} mention:

\begin{quote}
... we may infer that sexual desire for a man forms at least one aspect of third-sex sexuality. In a set of similes descriptive of the relative intensities of the sexualities of the three sexes, that of the third sex is viewed as most intense of all: a woman's \textit{veda} is compared to a dung fire, a man's to a forest fire, but the third sex's is compared to a burning city. Thus third-sex persons are not only sexual persons, but hyperlibidinous ones at that.
\end{quote}

The word \textit{napuṁsaka} has been the subject of much debate within the Jain Order, resulting over time in changes in meaning and use and the definition of sub-categories. The word in the canonical texts seems to have referred only to males who were cross-dressing and had a female gender-expression, who are identified by the way they dress, their behavior and sexual orientation. Because they looked like a woman, they were also assumed to have sexual desire for men. Because of this characterization, the \textit{napuṁsaka} can also be an object of lust for celibate monks and this would have been seen as problematic for ordination as a male monastic. Part of the discussion was also fuelled by the nakedness of the Jain monks and therefore their physical male appearance as well as their behavior. As celibate monks, same-sex relations and the possibility of same-sex attraction were issues. The public perception, and the fear thereof, was of utmost importance for the livelihood of the Jain Order. 

We also see a shift in the discussion over time regarding the abilities of a \textit{napuṁsaka}, or at least some sub-categories thereof, to attain enlightenment or to ordain. The \textit{Śvētāmbara} in their later \textit{Bhāgavatī Sūtra}\footnote{\textit{Bhāgavatī Sūtra} 4.1–2.} even define a fourth type, namely the \textit{puruṣanapuṁsaka} (male \textit{napuṁsaka}, possibly a \textit{napuṁsaka} who appeared the same as other men).\footnote{See \cite{zwilling} for more details.} Lacking any of the external characteristics of a \textit{napuṁsaka}, the only characteristic left to define them as such must have been their sexuality (i.e. attraction to men).

During the period of the commentarial literature the sexuality of the \textit{napuṁsaka} was redefined as being of a more bisexual orientation. Leonard Zwilling and Michael Sweet\footnote{\cite{zwilling} pages 371–374.} believe that this new definition was not so much driven by actual observations of the behavior of \textit{napuṁsakas} but rather by theoretical discussion. This bisexual orientation was not conceived of as a separate orientation, but as possessing the sexuality of \textbf{both} males and females. This is a change from the canonical literature, where the sexuality of a \textit{napuṁsaka} was characterized as female only.

During the period in which the commentarial literature was established the terms male and female \textit{napuṁsaka} were redefined further. The female \textit{napuṁsaka} was the old category defined in the Canon of which the \textit{klība} and \textit{paṇḍaka} are sub-categories, and the male \textit{napuṁsaka} was the aforementioned \textit{puruṣanapuṁsaka}. The female \textit{napuṁsaka} seems to act as a female partner only (i.e. be acted upon), while the male \textit{napuṁsaka} acts in both ways. So here male and female sexuality are no longer just defined as the sexual desire to have sex with a female and male respectively, but also in terms of the role taken in intercourse.\footnote{\textit{Niśītha Sūtra} 3507.} The hyperlibidinous nature of the \textit{napuṁsaka} was ascribed to the bisexual character of his sexuality. As we have seen above, from around the same time that the commentarial literature was established, the `third sex' became identified with their religious sexual role of providing alternative techniques of sexual gratification as in the \textit{Kāmasūtra}. It is therefore not surprising that they came to be seen as hyperlibidinous in nature.

It is interesting to note that throughout this discussion the \textit{napuṁsaka} and its sub-categories always refer to males who are somehow blocked in terms of exercising their male sexuality in one way or another owing to their performance of some unvirtuous act (\textit{karma}) in a previous life. Females who did not conform to gender-expectations were not considered in the class of \textit{napuṁsakas} or are only very rarely mentioned, without much explanation as to their nature. 

\subsection{Jain Monastic Ordination}
In the formative years of the Jain Order, the rules for ordination were still rather simple. Only the \textit{klība}, the \textit{paṇḍaka} and ill people were not allowed to ordain. Of the two Jain sects after the schism, the \textit{Digambara} maintained nakedness and eligibility to ordain as a monk was quite straightforward: one had to be a man without genital defects and virile, except for when that man was overly libidinous. 

For the \textit{Śvētāmbara}, who wore a cloth, the matter was far more complex and they devised an intricate system of people suitable for ordination, whereby the \textit{napuṁsaka} was divided into sixteen types.\footnote{See \textit{Bhagavatī} 5166–67.} Over time, the ban against ordination of \textit{napuṁsakas} was relaxed, first based on practical grounds like a known and well-behaved candidate, and later an exception was made for those who were able to control their sexuality. One of the main grounds why certain \textit{napuṁsakas} were denied ordination was their perceived hyperlibidinous-ness, which would render them incapable of keeping their vows of celibacy and made them unfit to live in either the monks' or nuns' communities. Only ten of the sixteen were not allowed to be ordained because they were regarded as uncontrollable in their passions. Among these were the original two categories of \textit{klība} and \textit{paṇḍaka}. The aforementioned \textit{puruṣanapuṁsaka} was allowed to ordain, presumably because they could not potentially evoke a monk's lust. Since external appearance was no longer a clear indication of being a \textit{napuṁsaka}, the candidate for ordination had to be questioned. 

By the 17\textsuperscript{th} Century CE, this rule on ordination had been nearly abolished. So we have seen a radical shift from total non-acceptance to almost total acceptance of \textit{napuṁsakas} in the Jain Order over time.\footnote{\cite{zwilling} reference \textit{Yuktiprabodha} in footnote 80.}
